---
title:  "Tensors"
author: Sudeep Sahadevan
date: "09/17/2015"
output:
 html_document:
  toc: true
  theme: cerulean
  highlight: tango
  
 pdf_document:
  toc: true
  theme: united
  highlight: tango  
---

###### A set of functions to do tensor analysis and higher order (genaralized) svd (hosvd|hogsvd)

## **get.tensor**

##### Description
Given a list of matrices and mode for the tensor, construct a tensor object of "atensor" class.

##### Parameters
* mat.list : a list of matrices
* mode : Tensor mode (Currently supports mode 1, 2 and 3) 

##### Usage
```{r eval=FALSE}
get.tensor(mat.list,mode=3)
```
##### Return
An object of "atensor" class which has the following attributes:

* ndim : numeric vector, the first two variables are the dimensions of input matrices, third variable, number of input matrices
* mode: A numeric variable, mode of the tensor
* dimnames: A list, row and column names of input matrices
* data: An array object, to hold the data

## **tensor.unfold**

##### Description
Given a tensor and the mode to unfold, unfold the tensor

##### Parameters
* tensor: an object of "atensor" class
* mode: the mode to unfold the tensor, three modes are supported at present, 1 column wise, 2 row wise, 3 tube wise
* verbose: boolean, controls the verbosity level 

##### Usage
```{r, eval=FALSE}
tensor.unfold(tensor,mode=1,verbose=TRUE)
```
##### Return
A matrix

## **tensor.fold**

##### Description

Given a matrix, the mode of the unfolded matrix and dimensions of the tensor, fold the matrix into  a tensor.

##### Parameters
* mat: input matrix
* mode: mode of the matrix, 1 for column, 2 for row and 3 for tube
* nmode: total modes of the tensor
* nrow: number of row elements for the tensor
* ncol: number of column elements for the tensor
* ntube: number of tube elements
* namedim: a list of row and column names for the input matrices (optional) 
* verbose: boolean, controls the verbosity level

##### Usage
```{r,eval=FALSE}
tensor.fold(mat,mode=1,nmode=3,n.row,n.col,n.tube,namedim=NULL,verbose=TRUE)
```
##### Return
An object of "atensor" class which has the following attributes:

* ndim : numeric vector, the first two variables are the dimensions of input matrices, third variable, number of input matrices
* mode: A numeric variable, mode of the tensor
* dimnames: A list, row and column names of input matrices
* data: An array object, to hold the data

## **get.fnorm**

##### Description

Given an object of class "atensor", "array", "matrix" or "data.frame" calculate the Frobenius norm.
For a matrix A, Frobenius norm is defined as $\parallel{A}\parallel_{F}=\sqrt[]{\sum_{j=1}^{M}\sum_{i=1}^{N}|a_{i\,j}|^2}$

##### Parameters
* obj: an object of "atensor" class OR "array" OR "matrix" OR "data.frame"

##### Usage
```{r,eval=FALSE}
get.fnorm(obj)
```

##### Return

A scalar value

## **ho.svd**

##### Description

Given an tensor, compute the higher order singular value decomposition (hosvd).

hosvd explanation: Tamara G. Kolda and Brett W. Bader (2009). Tensor Decompositions and Applications. SIAM Rev., 51(3), 455–500.

[Article DOI](http://epubs.siam.org/doi/abs/10.1137/07070111X)

[Pdf link](http://www.cs.cmu.edu/~christos/courses/826-resources/PAPERS+BOOK/Kolda-Bader-SAND2007-6702.pdf)

##### Parameters
* tensor: an object of "atensor" class
* ranks: ranks for decomposing the tensor, either a scalar or a vector satisfying the condition `length(ranks)==tensor@mode`, but this value cannot be NULL

##### Usage
```{r,eval=FALSE}
ho.svd(tensor,ranks = 10,verbose = TRUE)
```

##### Return

An list with the following elements:

* t.svd : tensor decomposition
* t.est : tensor re-estimated from the decomposition (t.svd)
* resid : residual, Frobenius norm of the difference between input tensor and the re-estimated t.est 

## **tensor.list.mult**

##### Description

Given a list of matrices and a tensor, unroll the tensor, multiply with the list and roll it back to  a tensor

##### Parameters
* tensor: an object of "atensor" class
* mat.list: a list of matrices
* mode.list: an integer vector with `length(list.mode) == length(matlist)` and the values  denote the mode in which the tensor must be unrolled
* verbose: A boolean controlling the verbosity level

##### Usage

```{r,eval=FALSE}
tensor.list.mult(tensor, mat.list, mode.list, verbose = TRUE)
```

##### Return

An object of "atensor" class which has the following attributes:

* ndim : numeric vector, the first two variables are the dimensions of input matrices, third variable, number of input matrices
* mode: A numeric variable, mode of the tensor
* dimnames: A list, row and column names of input matrices
* data: An array object, to hold the data

## **compute.hogsvd**

##### <a name="chogsvd"></a> Description

Given a list of matrices, compute the higher order generalized singular value decomposition as proposed by the publication:

Ponnapalli SP, Saunders MA, Van Loan CF, Alter O (2011). A Higher-Order Generalized Singular Value Decomposition for Comparison of Global mRNA Expression from Multiple Organisms. PLoS ONE 6(12): e28072

[Article DOI: 10.1371/journal.pone.0028072](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0028072)

and improvements for using similarity matrices for hogsvd were proposed by:

Xiao X, Moreno-Moral A, Rotival M, Bottolo L, Petretto E (2014) Multi-tissue Analysis of Co-expression Networks by Higher-Order Generalized Singular Value Decomposition Identifies Functionally Coherent Transcriptional Modules. PLoS Genet 10(1): e1004006

[Article DOI: 10.1371/journal.pgen.1004006](http://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1004006)

NOTE: #' NOTE: Moreover, ($W$ OR $S$ here)  is not guaranteed to be non-defective and have a full set of real eigenvalues and eigenvectors. However, even in the absence of an exact common decomposition, the real part of the complex eigenvectors can be used to derive a low rank approximation of the common subspace and extract common and differential covariance structures from the data.

HOGSVD of a list of N matrices are defined as:

$D_{1}\, =\, U_{1}\Sigma_{1}V^{T},$

$D_{2}\, =\, U_{2}\Sigma_{2}V^{T},$

$\vdots$

$D_{N}\, =\, U_{N}\Sigma_{N}V^{T},$

In case, the input list contains expression value matrices , the matrix $A$ is calculated as $A\, =\, D^{T}D$.

See [compute.right.singular](#crs) function for details on calculating the right singular eigensystem for N matrices. Given
$V$, the left singular system calculation is as follows:

$VB_{i}^{T}\, =\, D_{i}^{T},$

$B_{i}\, \equiv\, (b_{i,1},b_{i,2}...b{i,n}),\, i\, =\, 1,2...N$

$\sigma_{i,k}\, =\, \parallel{b_{i,k}}\parallel,$

$\Sigma_{i}\, =\, diag(\sigma_{i,k}),$

$B_{i}\, =\, U_{i}\Sigma_{i}$.

##### Parameters

* matlist: a list of matrices, the matrices could be $n\times{m}$ expression matrices ($D$) or $n\times{n}$ similarity matrices ($A$)
* crossp: Boolean, if `TRUE`, calculates matrix crossproduct $A\, =\, D^{T}D$ 
* pinv: Boolean, if `TRUE`, calculates Moore–Penrose pseudoinverse ($A^{+}$) of the matrix, else calculates the general inverse ($A^{-1}$) of a matrix

##### Usage
```{r,eval=FALSE}
compute.hogsvd(matlist, crossp = FALSE, pinv = TRUE)
```

##### Return

An object of "hogsvd" class with the following attributes:

* Ulist: a list of left singular matrices, $U_{i}$,  where `length(Ulist)==length(matlist)`
* Dlist: a list of diagonal vectors from the decomposition of matlist, $\Sigma_{i}$
* V: the generalized right singular matrix computed from the input matlist, $V$
* Vd: a  vector of diagonal elements from hogsvd calculation, $\lambda_{k}$

## **compute.right.singular**

##### <a name="crs"></a>Description

Given a list of matrices, compute the generalized right singular eigensystem for the matrices.
This function is a part of the function [compute.hogsvd](#chogsvd). Refer [compute.hogsvd](#chogsvd) function for article details.

Right singular eigensystem for a list of $N$ matrices is calculated as:

$S\,\equiv\,\frac{1}{N(N-1)}\sum_{i=1}^{N}\sum_{i>j}^{N}(A_{i}A_{j}^{-1}+A_{j}A_{i}^{-1}),$

$SV\, =\, V\Lambda,$

$V\, \equiv\, (v_{1},v_{2}...v_{n}),\, \Lambda\, =\, diag(\lambda_{k})$ with

$\parallel{v_{k}}\parallel=1$.

##### Parameters

* matlist: a list of matrices, the matrices could be $n\times{m}$ expression matrces or $n\times{n}$ similarity matrices
* pinv: Boolean, if `TRUE`, calculates Moore–Penrose pseudoinverse ($A^{+}$) of the matrix, else calculates the general inverse ($A^{-1}$) of a matrix

##### Usage

```{r,eval=FALSE}
compute.right.singular(matlist, pinv = FALSE)
```

##### Return
Returns the results from the function `eigen`, a list with attributes `values` ($\lambda_{k}$) and `vectors` $V$.
